---
title: "Sorting Algorithms in R"
author: "Rafael Carlos Abbariao"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    self_contained: false
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
---

I’ve just read “Algorithms to Live By: The Computer Science of Human Decisions” and there’s a neat sorting chapter that made these concepts a little more tangible and I will try to show some examples of these sorting methods in R.

Sorting is the backbone of data analysis and there are a few methods discussed in the aforementioned book:

- Bubble Sort
- Insertion Sort
- Merge Sort

As a reminder, we have some regular sorting functions in base R.

```{r}
set.seed(123)
n <- sample(1:100, 10)

sort(n) # sorts the vector in ascending order (or descending with decreasing = TRUE) order

rank(n) # returns the sample ranks of the values in a vector

order(n) # returns the indicces of the vector in a sorted order
```

But onto the sorting algorithms mentioned….

## Bubble Sort

Bubble sort is the most “brute force” sorting method of the few listed. Appropriately enough, the analogy provided in the book is ordering books on a bookshelf.

Given a number of disorganized books, we compare the first two adjacent book titles and reordering the “larger” element to the right (if sorting in ascending or alphabetical order). This is repeated sequentially until the end of the series and until all elements are in ascending order.

For example:

Step 0: Moby Dick -> Algorithms to Live By -> Holy Bible
Step 1: Algorithms to Live By -> Moby Dick -> Holy Bible
Step 2: Algorithms to Live By -> Holy Bible -> Moby Dick

```{r}
bubbleSort <- function(n) {
  len <- length(n)
  for (i in 1:(len - 1)) {
    for (j in 1:(len - i)) {
      if (n[j] > n[j + 1]) {
        # Swap elements
        temp <- n[j]
        n[j] <- n[j + 1]
        n[j + 1] <- temp
      }
    }
  }
  return(n)
}
```


## Insertion Sort

Within the same bookshelf OCD analogy, we begin with an empty shelf. After arbitrarily sorting one book on the shelf, we then compare the next one with the “sorted” book, and so on until all are organized.

```{r}
insertionSort <- function(n) {
  for (i in 2:length(n)) {
    temp = n[i]  # set the first unsorted element for comparison
    j = i - 1
    while(n[j] > temp && j > 0) {
      n[j + 1] = n[j]
      j = j - 1
    }
    n[j + 1] = temp
  }
  return(n)
}
```

## Merge Sort

The merge sort can also be referenced as the divide and conquer method, not the fascist, age of empires kind but the clean up, everyone do your part kind.

Back to the bookshelf thing… merge sort would involve multiple people (imagine having friends) with an equal number of random books to start. Each person organizes their own stack, finds a partner to merge their stack in order and so on until all stacks are merged and sorted.

```{r}
# function to merge sorted arrays (to be used in main sorting function)
merge <- function(x, y) {
  temp = numeric(length(x) + length(y))  # initialize temporary array
  startx = 1
  starty = 1
  j = 1  # start of temp array
  
  for (j in 1:length(temp)) {
    if ((startx <= length(x) && x[startx] < y[starty]) || starty > length(y)) {
      temp[j] = x[startx]
      startx = startx + 1
    } else {
      temp[j] = y[starty]
      starty = starty + 1
    }
  }
  temp
}

mergeSort <- function(n) {
  if (length(n) > 1) {
    # find mid point of array
    mid <- ceiling(length(n)/2)
    x <- mergeSort(n[1:mid])
    y <- mergeSort(n[(mid+1):length(n)])
    merge(x, y)
  } else {
    n
  }
}
```

In comparing these sorting methods we find that …

```{r}
library(ggplot2)
library(microbenchmark)

set.seed(123)

n <- sample(1:1000, 1000, replace = TRUE)

comparison <- microbenchmark(
   bubbleSort = bubbleSort(n),
   insertionSort = insertionSort(n),
   mergeSort = mergeSort(n),
   times = 100
)

autoplot(comparison) +
  labs(title = "Sorting Algorithms Comparison",
       x = "Time (in nanoseconds)",
       y = "Algorithms") +
  theme_minimal()
```

… merge sort is the best option, followed by insertion sort and close after, bubble sort. This brings us into the world of space and time complexity of algorithms.